A thermodynamic view of dusty protoplanetary disks

### 2017-09-11 paper.tex
revising in response to report

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas friction introduces phenomena
that are absent in pure gas disks. Important examples include the
streaming instability 
\citep[SI,][]{youdin05a,youdin07b,johansen07}, secular gravitational
instabilities \citep[SGI,][]{ward00,youdin11,michikoshi12,takahashi14,latter17} and 
Kelvin-Helmholtz instabilities
\citep{goldreich1973, chiang08,barranco09,lee10}.

@@

This paper is organized as follows. 
In \S\ref{grow_osc} 
we describe a key insight of our study, that pressure (i.e.\ ``PdV") work explains the basis of several instabilities, including SI and VSI, in terms of
pressure-density phase lags.    
We develop our formalism in \S\ref{setup} by transforming the
two fluid equations of a tightly-coupled dusty gas into single fluid
hydrodynamic equations with a special cooling function.  Here we also
define the effective entropy and buoyancy of our model fluid. 
%We derive equilibrium solutions in \S\ref{eqm}, which we use in our
%explicit stability calculations.
In \S\ref{limits} we discuss general stability properties of dusty
disks. We then analyze explicit disk models 
\S\ref{linear_problem}, were we show that radial dust edges are unstable,
and revisit the SGI and SI in the tight 
coupling regime. In \S\ref{results} we 
extend the VSI to dusty disks. We   
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

@@

\section{Growing oscillations by doing work}\label{grow_osc}
A general result in fluid dynamics is that work is done whenever
oscillations in the pressure and density of the fluid are not in
phase. If  the average work done is positive, then oscillation
amplitudes grow: the work done allows the system to `overshoot' beyond 
the amplitude of preceding oscillations.

@@

\input{model}
\input{linear}
\input{result}
\input{summary}
\acknowledgements
We thank P. Loren-Aguilar for initial discussions that motivated this study. We also thank 
C. Baruteau, R. Dong, J. Fung, S. Inutsuka, W. Lyra, and M. Pessah 
for comments during the course of this project. This work is supported by the Theoretical Institute for Advanced Research in Astrophysics 
(TIARA) based in Academia Sinica Institute of Astronomy and Astrophysics (ASIAA) and NASA Astrophysics Theory Program grant NNX17AK59G.

>>>>

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas friction introduces phenomena
that are absent in pure gas disks. Important examples include the
streaming instability 
\citep[SI,][]{youdin05a,youdin07b,johansen07}, secular gravitational
instabilities \citep[SGI,][]{ward00,youdin11,michikoshi12,takahashi14,taka16,latter17} and 
Kelvin-Helmholtz instabilities
\citep{goldreich1973, chiang08,barranco09,lee10}.

@@

This paper is organized as follows.  
In \S\ref{grow_osc} we describe a key insight of our study, that
pressure (i.e.\ ``PdV")  work {\bf can} explain the basis of several
instabilities, including SI and VSI, in terms of 
pressure-density phase lags. {\bf This motivates us to seek an 
  analogy between dust-gas mixtures and pure gas.}    
We develop our formalism in \S\ref{setup} by transforming the
two fluid equations of a tightly-coupled dusty gas into single fluid
hydrodynamic equations with a special cooling function.  Here we also
define the effective entropy and buoyancy of our model fluid. 
In \S\ref{limits} we discuss general stability properties of dusty
disks. We then analyze explicit disk models 
\S\ref{linear_problem}, were we show that radial dust edges are unstable,
and revisit the SGI and SI in the tight 
coupling regime. In \S\ref{results} we 
extend the VSI to dusty disks. We   
discuss future applications of the dusty/adiabatic gas equivalence in 
\S\ref{discussion} before summarizing in \S\ref{summary}.

@@

\section{Growing oscillations by doing work}\label{grow_osc}
{\bf
In this section we provide physical arguments to highlight the
similarity between single phase fluids and dust-gas mixtures. It is
well-known that in the limit of perfect coupling the addition of dust
increases the gas inertia but not pressure. One can then regard the
dusty gas as a single mixture with a reduced sound-speed. Here, we
argue that when dust-gas coupling is imperfect, there exists another
similarity with pure gas related the phase of pressure and density
evolution. Thus it is still possible to regard
partially-coupled dusty-gas as a single fluid. The mathematical
description of this section is given in \S\ref{dust_work}.   
}
%dust_work

A general result in fluid dynamics is that work is done whenever
oscillations in the pressure and density of the fluid are not in
phase. If  the average work done is positive, then oscillation
amplitudes grow: the work done allows the system to `overshoot' beyond 
the amplitude of preceding oscillations.

@@

\input{model}
\input{linear}
\input{result}
\input{summary}
\acknowledgements
We thank P. Loren-Aguilar for initial discussions that motivated this study. We also thank 
C. Baruteau, R. Dong, J. Fung, S. Inutsuka, W. Lyra, M. Pessah, and O. Umurhan 
for comments during the course of this project. This work is supported
by the Theoretical Institute for Advanced Research in Astrophysics  
(TIARA) based in Academia Sinica Institute of Astronomy and
Astrophysics (ASIAA), NASA Astrophysics Theory Program grant
NNX17AK59G, and  
the Steward Theory Fellowship at the University of Arizona.

### 2017-09-11 result.tex
revising in response to report

We now consider a radially-varying dust-to-gas ratio. Specifically we
let $\epsilon_0\propto r^{-1}$ and $H_\epsilon \propto \Hg$
(cf. constant values in the previous calculations). Then
\begin{align*} 
  \frac{\p\epsilon}{\p r} =
  \left(\frac{z^2}{H_\epsilon^2}\frac{d\ln{\Hg}}{dr} -
    \frac{1}{r}\right)\epsilon.   
\end{align*}
Here we fix $Z=0.01$, $\Hd=0.8\Hg$, and $k_x\Hg=30$.

Fig. \ref{compare_vshear_varHd} shows that a radially-varying
dust-to-gas ratio increases the magnitude of the vertical shear rate
away from the midplane (see also Eq. \ref{vshear2}). Thus we find
higher VSI growth rates, as shown in Fig. \ref{vsi_dust_varHd}. 
Surfaces modes are more effectively enhanced by the additional vertical 
shear induced by $\p_r\epsilon$. However, all growth rates remain
$O(\hgas\OmK)$. Importantly, the increase in the growth rate of the
fundamental mode is small. We do not expect the radial
dependence in $\epsilon$ to significantly affect the VSI in a dusty
disk.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_varHd} 
  \caption{Unstable VSI modes in the disk models of
    Fig. \protect\ref{compare_vshear_varHd}.  Black (blue) diamonds
    are mode frequencies for a disk with radially uniform (varying)
    dust-to-gas ratio. 
    \label{vsi_dust_varHd}
    }
\end{figure}

@@

In \S\ref{iso_perfect} we found that for strictly isothermal
disks, a vertically-uniform dust-to-gas ratio can be unstable
if $d\epsilon/dr\neq0$.  
To demonstrate this numerically we consider the
same disk as above but with $q=0$ and choose $H_\mathrm{d}$ such
that $H_\epsilon=10^3\Hg$. Then vertical shear $\p_z\Omega$ is 
attributed to the radially-varying dust-to-gas ratio (see
Eq. \ref{vshear2}).

We show unstable modes for this disk in
Fig. \ref{vert_mixed_modes}. In order to obtain appreciable growth   
rates, we consider super-Solar metalicity $Z=0.03$ and high
radial wavenumber $k_x\Hg=1800$. As expected from the discussion in
\S\ref{iso_perfect}, the disk admits purely growing modes with
$\omega=0$. This is distinct from the growing oscillations associated
with classic VSI discussed above.

@@

>>>>

We now consider a radially-varying dust-to-gas ratio. Specifically we
let $\epsilon_0\propto r^{-1}$ and $H_\epsilon \propto \Hg$
(cf. constant values in the previous calculations). Then
\begin{align*} 
  \frac{\p\epsilon}{\p r} =
  \left(\frac{z^2}{H_\epsilon^2}\frac{d\ln{\Hg}}{dr} -
    \frac{1}{r}\right)\epsilon.   
\end{align*}
Here we fix $Z=0.01$, $\Hd=0.8\Hg$.

Fig. \ref{compare_vshear_varHd} shows that a radially-varying
dust-to-gas ratio increases the magnitude of the vertical shear rate
away from the midplane (see also Eq. \ref{vshear2}). Thus we {\bf
  typically} find higher VSI growth rates, as shown in
Fig. \ref{vsi_dust_varHd}. 
Surfaces modes, {\bf appearing at high $k_x$}, are more effectively
enhanced by the additional vertical shear induced by
$\p_r\epsilon$. {\bf We find that low frequency body modes with small
  $k_x$ are more affected by radial variations in the 
  dust-to-gas ratio.}

However, all growth rates remain
$O(\hgas\OmK)$. Importantly, the increase in the growth rate of the
fundamental mode is small. We do not expect the radial
dependence in $\epsilon$ to significantly affect the VSI in a dusty
disk.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_eigenvals_varHd} 
  \caption{Unstable VSI modes in the disk models of
    Fig. \protect\ref{compare_vshear_varHd}. {\bf Diamonds (asterisks) 
    are mode frequencies for a disk with radially uniform (varying)
    dust-to-gas ratio. The radial wavenumbers are $k_xH_g = 10$
    (green),$k_xH_g = 30$ (black) and $k_xH_g = 100$ (orange). }
    \label{vsi_dust_varHd}
    }
\end{figure}

@@

In \S\ref{iso_perfect} we found that for strictly isothermal
disks, a vertically-uniform dust-to-gas ratio can be unstable
if $d\epsilon/dr\neq0$.   
{\bf
To demonstrate this numerically we set $q=0$ and $H_\mathrm{d}$ such
that $H_\epsilon=10^3\Hg$. We let $\rhod/\rhog \propto r^{-d}$
with different power indices $d$. The metalicity is fixed to $Z=0.03$.  
In these cases vertical shear $\p_z\Omega$ is 
attributed to the radially-varying dust-to-gas ratio (see
Eq. \ref{vshear2}).    
}

{\bf We show unstable modes in 
Fig. \ref{vert_mixed_modes}.} 
%In order to obtain appreciable growth   
%rates, we consider super-Solar metalicity $Z=0.03$ and high
%radial wavenumber $k_x\Hg=1800$. 
As expected from the discussion in
\S\ref{iso_perfect}, the disk admits purely growing modes with
$\omega=0$. This is distinct from the growing oscillations associated
with classic VSI discussed above.

{\bf We find in smooth disks large 
  $k_x$ is needed for appreciable growth rates, e.g. $k_xH_g=1800$
  with $\rhod/\rhog\propto r^{-1}$. 
  Local shearing box simulations \citep[e.g.][]{bai10b,yang16} may 
  be required to study the non-linear evolution of such short 
  wavelengths. Alternatively, as shown in Fig. \ref{vert_mixed_modes},
  a rapidly varying dust-to-gas ratio permits dynamical instability at
  longer radial wavelengths. 
}

@@

%{\bf
%  \subsection{Implications for protoplanetary disks}
%dust settling, dust-induced buoyancy 
%The above calculations 
%radial d/g -> radial temp profile 
%}

### 2017-09-11 summary.tex
revising in response to report

@@

For the VSI we find dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed where the dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}. Since the dust-induced buoyancy forces
increase away from the midplane, we find dust-loading can stabilize
`surface modes'  of the VSI, that would otherwise have the largest
growth rates.

>>>>

%pros and cons 
%energy analogy not restricted to term vel approx 
%spurious modes

@@

For the VSI we find dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed where the dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}. 
Since the dust-induced buoyancy forces
increase away from the midplane, we find dust-loading can stabilize
`surface modes'  of the VSI, that would otherwise have the largest
growth rates. {\bf We also showed that radial variations in
  $\rhod/\rhog$ translates to an effectively non-uniform radial temperature
  profile, and the resulting vertical shear can lead to instability,
  even if the true sound-speed is constant. 
}

%implications for realistic disks
{\bf In a realistic disk, dust particles settle on a timescale $
  t_\mathrm{settle} \sim 1/\tstop\OmK^2$ \citep{takeuchi02}, compared 
  with typical VSI growth timescales, $t_\mathrm{grow}\sim 
  1/h_\mathrm{g}\OmK$. This suggest that particles with $\tstop\OmK\lesssim 
  h_g$ cannot settle against the VSI. On the 
  other hand, larger particles with $\tstop\OmK\gtrsim h_g$ should
  settle to form a dusty midplane. In fact, our calculations suggest that
  settling would stabilize the disk against VSI, and allow
  further settling. The result may be a quiet, dusty midplane \citep[unless non-axisymmetric
  instabilities develop][]{chiang08} with VSI-turbulent gaseous
  atmospheres. 
}
